---
title: 'tidyverse: World Population Solutions'
author: "Type your name here!"
date: "`r format(Sys.time(), '%a %d %b %Y (%H:%M:%S)')`"
output:
  html_document:
    highlight: tango
    theme: united
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
      smooth_scroll: no
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## World Population Data

We will use a subset of the United Nations world population data[^UNdata]. 
The subset has 5 yearly population estimates from 1950 until 2015 by age groups for
both males and females. The variables in the dataset are as follows.

- `continent`: Continent
- `Location`: Country name
- `Time`: Year
- `AgeGrp`: Age group (e.g. 15-19, 20-14)
- `AgeGrpStart`: initial age of the age group
- `PopMale`: Total male population (thousands)
- `PopFemale`: Total female population (thousands)

## Exercise Questions

We want to answer the following questions with data:

1. How did the Portuguese population change between 1950 and 2015?
2. What was the total world population by sex in 1950 and 2015?
3. What was the total population by continent and sex in 1950 and 2015?
4. What was the total population by country and sex in 1950 and 2015?
5. How did the total population of Portugal change between 1950 and 2015?
6. Which countries had the highest and lowest proportion of females in 1950 
and 2015?
7. Discuss how the proportion of Portuguese females changed between 1950 and 2015?


## Read in the data

First we will load the `tidyverse` packages using the `library()` fucntion. Then
we will read the data from file `World-Population.csv` into object called `unOrig`. 

```{r freadData}
# Type answer here.
library(tidyverse)
library(here)
unOrig <- read_csv(here("World-Population.csv"))

unOrig
glimpse(unOrig)
```

### Pre-processing the Data

Start by preparing the data for analysis in a new object called `un`:

1. Rename `continent` to `Continent`, `Location` to `Country` and `Time` to `Year`. 
2. Create variable `PopF` as 1000 times `PopFemale`. 
3. Create variable `PopM` as 1000 times `PopMale`. 
4. Create variable `PopT` as total of males and females.
5. Drop variables `PopFemale` and `PopMale`.
6. Order the data by continent, country, year and age group
7. Look at the data (by typing `un` and using the `glimpse()` function).

**Important** Use the pipe operator (`%>%`) for all the exercises.

```{r preprocess}
# Type answer here.
un <- unOrig %>% 
  rename(Continent=continent, Country=Location, Year=Time) %>% 
  mutate(
    PopF = 1000*PopFemale,
    PopM = 1000*PopMale,
    PopT = PopF + PopM) %>% 
  arrange(Continent, Country, Year, AgeGrp)

un
```

**Are the age groups correctly ordered?** If not, why not? How can you correct this?

```{r ageGrpArrange}
un <- unOrig %>% 
  rename(Continent=continent, Country=Location, Year=Time) %>% 
  mutate(
    PopF = 1000*PopFemale,
    PopM = 1000*PopMale,
    PopT = PopF + PopM) %>% 
  arrange(Continent, Country, Year, AgeGrpStart)

un
```


## Exercise Answers

### Portugal: 1950 v 2017

> How did the Portuguese population change between 1950 and 2015?

1. Create a dataset called `popPT`
2. Filter to select Portugal and the years 1950 and 2015 (hint use `%in% c(...)`)
3. Arrange this dataset by Year and age group

**What do you notice about the variable `AgeGrp`?**

```{r PT19502017}
# Type answer here.
popPT <- un %>% 
  filter(Country=="Portugal", Year %in% c(1950, 2015)) %>% 
  arrange(Year, AgeGrpStart)

popPT
```

### World Population

> What is the total world population by year?

1. Create a new object called `wp` to assign the total world population by year. 
2. Create `TotF`, `TotM and `Total` as the total for females, males and both.
3. Arrange by year.
4. Look at the `wp`.

```{r WorldTotal}
# Type answer here.
wp <- un %>% 
  group_by(Year) %>% 
  summarise(TotF = sum(PopF), TotM = sum(PopM), Total = sum(PopT)) %>% 
  arrange(Year)
```

**Discuss the changes between 1950 and 2015.**

### Population by Continent

> What is the total population by continent and sex in 1950 and 2015?

1. Create a new object called `ct` to assign total population by continent and year. 
2. Only keep data for 1950 and 2015.
3. Create `TotF`, `TotM and `Total` as the total for females, males and both.
4. Arrange the data by Contient and Year
4. Look at the `ct`.


```{r ContinentTotal}
# Type answer here.
ct <- un %>% 
  filter(Year %in% c(1950, 2015)) %>% 
  group_by(Continent, Year) %>% 
  summarise(TotF = sum(PopF), TotM = sum(PopM), Total = sum(PopT)) %>% 
  arrange(Continent, Year)

ct
```

**What changes do you observe for each continent between 1950 and 2015? Which continent
has the largest change? Which has the smallest change?**

### Population by Country

> What is the total population by country and sex in 1950 and 2015?

1. Create a new object called `cy` to assign total population by country. 
2. Only keep data for 1950 and 2015.
3. Create `TotF`, `TotM and `Total` as the total for females, males and both. Keep 
the variable `Continent` in the `by` part.
5. Look at the `cy`.

```{r CountryTotal}
# Type answer here.
cy <- un %>% 
  filter(Year %in% c(1950, 2015)) %>% 
  group_by(Continent, Country, Year) %>% 
  summarise(TotF = sum(PopF), TotM = sum(PopM), Total = sum(PopT)) %>% 
  arrange(Continent, Country, Year)

cy
```


### Portugal: 1950 to 2015

> How did the total population of Portugal change between 1950 and 2015?

1. Select Portugal in the `cy` dataset.
2. Look at other countries to see what is happening.

```{r PTchange}
# Type answer here.
cy %>% filter(Country=="Portugal")
```


### High & Low Female Proportion

> Which countries had the highest and lowest proportion of females in 1950 
and 2015?

1. Start with dataset `cy` to create a new dataset called `pp`.
2. Calculate two new variables called `PropF` and `PropM` for the proportion of females and males, respectively.
3. Order `PropF` from the lowest to the highest and select the first 10 observations
for both 1950 and 2015.
4. Order `PropF` from the highest to lowest and select the first 10 observations
for both 1950 and 2015.



```{r proportion}
# Type answer here.
pp <- cy %>% 
  mutate(
    PropF = TotF/Total,
    PropM = TotM/Total
  ) %>% 
  group_by(Year) %>% 
  arrange(Year, PropF)

# Top 10 (lowest) for each year
pp %>% top_n(n=10)

# Last 10 (highest) for each year
pp %>% top_n(n=-10)
```


### Portuguese Females Over Time

> Discuss how the proportion of Portuguese females changed between 1950 and 2015?

1. Select Portugal from the `pp` dataset.

```{r ptChanges}
# Type answer here.
pp %>%  filter(Country=="Portugal")

```

[^UNdata]: Source: https://population.un.org/wpp/Download/Standard/CSV/
[Dataset: "WPP2017_PopulationByAgeSex_Medium.csv](https://population.un.org/wpp/DVD/Files/1_Indicators%20(Standard)/CSV_FILES/WPP2017_TotalPopulationBySex.csv)
